---
title: "Untitled"
---

```{r}
# load R packages
pacman::p_load(corrplot, ggstatsplot, tidyverse)
```

```{r}
# import data
wine <- read_csv("data/wine_quality.csv")
```

# 1. Building Correlation Matrix: pairs() method

## 1.1 Building a basic correlation matrix

```{r}
pairs(wine[,1:11])
```

The required input of pairs() can be a matrix or data frame. The code chunk used to create the scatterplot matrix is relatively simple. It uses the default pairs function. Columns 2 to 12 of wine dataframe is used to build the scatterplot matrix. The variables are: fixed acidity, volatile acidity, citric acid, residual sugar, chlorides, free sulfur dioxide, total sulfur dioxide, density, pH, sulphates and alcohol.

```{r}
pairs(wine[,2:12])
```

## 1.2 Drawing the corner

```{r}
pairs(wine[,2:12], upper.panel = NULL)
```

```{r}
pairs(wine[,2:12], lower.panel = NULL)
```

## 1.3 Including with correlation coefficients

To show the correlation coefficient of each pair of variables instead of a scatter plot,[*panel.cor*](https://www.rdocumentation.org/packages/xcms/versions/1.48.0/topics/panel.cor)function will be used. This will also show higher correlations in a larger font.

```{r}
#| warning: false
panel.cor <- function(x, y, digits=2, prefix="", cex.cor, ...) {
usr <- par("usr")
on.exit(par(usr))
par(usr = c(0, 1, 0, 1))
r <- abs(cor(x, y, use="complete.obs"))
txt <- format(c(r, 0.123456789), digits=digits)[1]
txt <- paste(prefix, txt, sep="")
if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
text(0.5, 0.5, txt, cex = cex.cor * (1 + r) / 2)
}

pairs(wine[,2:12], 
      upper.panel = panel.cor)
```

# 2. Visualising Correlation Matrix: ggcormat()

One of the major limitation of the correlation matrix is that the scatter plots appear very cluttered when the number of observations is relatively large (i.e. more than 500 observations). To over come this problem, **Corrgram** data visualisation technique suggested by D. J. Murdoch and E. D. Chow (1996) and Friendly, M (2002) and will be used.

The are at least three R packages provide function to plot corrgram, they are:

-   [corrgram](https://cran.r-project.org/web/packages/corrgram/)
-   [ellipse](https://cran.r-project.org/web/packages/ellipse/)
-   [corrplot](https://cran.r-project.org/web/packages/corrplot/)

On top that, some R package like ggstatsplot package also provides functions for building corrgram.

In this section, you will learn how to visualising correlation matrix by using [*ggcorrmat()*](https://indrajeetpatil.github.io/ggstatsplot/reference/ggcorrmat.html) of [ggstatsplot](https://indrajeetpatil.github.io/ggstatsplot/) package.

```{r}
ggstatsplot::ggcorrmat(
  data = wine, 
  cor.vars = 1:11)
```